---
title: "Changepoint Detection for Open Source Monthly Time Series"
author: "James Walden"
date: "11/22/2020"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(changepoint)
library(changepoint.np)
library(tidyverse)
```

# Example: single change in mean

We create a dataset of 100 randomly distributed data points, the first 50 with mean 0 and the second 50 with mean 5, creating a clear changepoint at 50.
```{r}
set.seed(1)
ocp <- data.frame(
  x = 1:100,
  y = c(rnorm(50,0,1), rnorm(50,5,1))
)
```

We first use `changepoint` to find the changepoint in the series at 50.
```{r}
one.cpt <- cpt.mean(ocp$y)
cpts(one.cpt)
```

We next use the nonparametric variant `changepoint.np` to find the changepoint in the series with the same result.
```{r}
one.cpt.np <- changepoint.np::cpt.np(ocp$y)
cpts(one.cpt.np)
```

# Example: two changes of mean

We create a dataset of 100 randomly distributed data points, with two changes in mean at 30 and 70.
```{r}
tcp <- data.frame(
  x = 1:100,
  y = c(rnorm(30, 2), rnorm(40, 0), rnorm(30, 1))
)
plot(tcp)
abline(v = c(30, 70), col="red")
```

We first use `changepoint`, finding only one changepoint in the series, the first one at 30.
```{r}
tcp.cpt <- cpt.mean(tcp$y)
cpts(tcp.cpt)
```

We next use `changepoint.np` which finds both changepoints at 30 and 70.
```{r}
tcp.cpt.np <- changepoint.np::cpt.np(tcp$y)
cpts(tcp.cpt.np)
```

# Example: no changes in mean

We create a dataset of 100 randomly distributed data points, with no changes in mean.
```{r}
zcp <- data.frame(
  x = 1:100,
  y = c(rnorm(100,0,1))
)
plot(zcp)
```

We first use `changepoint`, which returns an empty vector with no changepoints.
```{r}
zcp.cpt <- cpt.mean(zcp$y)
cpts(zcp.cpt)
```
```{r}
length(cpts(zcp.cpt))
```

We next use `changepoint.np` which finds the same result.
```{r}
zcp.cpt.np <- changepoint.np::cpt.np(zcp$y)
cpts(zcp.cpt.np)
```

# Finding changepoints in OpenSSL

I'm using OpenSSL data from my MSR'20 paper, which includes time series data from project inception to the end of 2019.

I select data from 2012-2016 to analyze and call it commits for ease of typing.

```{r load_openssl_data}
openssl_commits <- read_csv('openssl-commits-per-month.csv', col_types="Di")
openssl_nearhb_commits <- openssl_commits %>% filter(Date >= as.Date('2012-01-01') & Date <= as.Date('2016-12-31'))
commits <- openssl_nearhb_commits
commits
```
```{r, echo=FALSE, fig.height=4, fig.width=8}
g <- ggplot(commits, aes(x=Date, y=ncommits)) + geom_point() + geom_line(group=1) + xlab("Month") + ylab("Number of Commits")
g
```

We first try changepoint.
```{r}
openssl.cpt <- cpt.mean(commits$ncommits)
changepoints <- cpts(openssl.cpt)
changepoints
```
```{r}
openssl.cp.date <-  commits %>% slice(changepoints) %>% select(Date) %>% pull()
commits %>% slice(changepoints)
```
```{r}
g + geom_vline(xintercept=openssl.cp.date, color="blue", linetype="dashed") + 
    ggtitle("changepoint package changepoints in OpenSSL commits 2012-2016")
```

We next use the nonparametric variant `changepoint.np` which finds two changepoints.
```{r}
openssl.cpt.np <- cpt.np(commits$ncommits)
changepoints.np <- cpts(openssl.cpt.np)
changepoints.np
```

```{r}
openssl.cp.dates <- commits %>% slice(changepoints.np) %>% select(Date) %>% pull()
commits %>% slice(changepoints.np)
```
```{r}
g + geom_vline(xintercept=openssl.cp.dates, color="red", linetype="dashed") + 
    ggtitle("changepoint.np changepoints in OpenSSL commits 2012-2016")
```

