---
title: "Changepoint EDA"
author: "James Walden"
date: "12/4/2020"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(changepoint.np)
library(fs)
library(lubridate)
```

```{r data, include=FALSE}
data_dir <- '../data'
commits_dir <- path(data_dir, 'commits')
ncommits_path <- path(data_dir, 'ncommits.csv')
nchangepoints_path <- path(data_dir, 'nchangepoints.csv')

pcommits <- read_csv(ncommits_path, col_names=c('nmonths','project'), col_types="ic")
pchangepoints <- read_csv(nchangepoints_path, col_names=c('nchangepoints','project'), col_types="ic")
```

```{r data_functions, include=FALSE}
read_commit_ts <- function(commits_path) {
    colnames <- c('year_month','ncommits','timestamp')
    commits_ts <- read_csv(file=commits_path, col_names=colnames, col_types='cii')
    commits_ts <- commits_ts %>%
        mutate(date=as_date(str_c(year_month, '-01')), timestamp=NULL) %>%
        complete(date=seq.Date(min(date), max(date), by="month"), fill=list(ncommits=0)) %>%
        mutate(year_month=str_c(year(date),str_pad(month(date),2,"left","0"),sep='-'),
               t=row_number(),
               project=path_ext_remove(path_file(commits_path))
              )
    commits_ts
}
```

```{r changepoint_functions}
find_changepoints <- function(commits_ts) {
    changes <- tibble(
                    row = integer(), 
                    date = Date(), 
                    pre_mean = double(),
                    post_mean = double(),
                    diff_means = double()
                    )

    if (nrow(commits_ts) > 1) {
        cnp <- cpt.np(commits_ts$ncommits, minseglen=2)
        changepoints <- cpts(cnp)
        nchangepoints <- length(changepoints)
    } else { # can't find change points with only one time period
        nchangepoints <- 0
    }

    if (nchangepoints > 0) {
        for (i in 1:nchangepoints) { 
            if (i == 1) {
                begin_row <- 1
            } else {
                begin_row <- changepoints[i-1]
            }
            if (i == nchangepoints) {
                end_row <- nrow(commits_ts)
            } else {
                end_row <- changepoints[i+1]
            }
            pre_change_mean <- commits_ts %>% 
                slice(begin_row:changepoints[i]) %>%
                summarize(mean(ncommits)) %>%
                pull()
            post_change_mean <- commits_ts %>% 
                slice(changepoints[i]:end_row) %>%
                summarize(mean(ncommits)) %>%
                pull()
            changes <- changes %>% add_row(row=changepoints[i],
            date=commits_ts %>% slice(changepoints[i]) %>% select(date) %>% pull(),
            pre_mean=pre_change_mean,
            post_mean=post_change_mean,
            diff_means=post_change_mean - pre_change_mean)
        }
    }
    changes
}
```


```{r graph_functions}
# Plot time series graph of a single project
plot_commit_ts <- function(ts) {
  ggplot(data=ts, aes(x=date,y=ncommits)) + geom_point() + geom_line() +
    xlab("Date") +
    ylab("Number of Commits")
}

# Plot time series graph of a single project with changepoints as verticle dashed lines
plot_commit_ts_with_cps <- function(ts) {
  cnp <- cpt.np(ts$ncommits, minseglen=2)
  changepoints <- cpts(cnp)
  nchangepoints <- length(changepoints)
  
  g <- ggplot(data=ts, aes(x=date,y=ncommits)) + geom_point() + geom_line() +
        xlab("Date") +
        ylab("Number of Commits")
  
  if (nchangepoints > 0) {
    for(cp in changepoints) {
      g <- g + geom_vline(xintercept=, color="blue", linetype="dashed") 
    }
  } else {
    g
  }
}

# Merge time series to use in plot_multiple_ts()
merge_commit_ts <- function(orig_ts, append_ts) {
  orig_ts <- orig_ts %>% relocate(project, t)
  append_ts <- append_ts %>% relocate(project, t)
  bind_rows(orig_ts, append_ts)
}

# Plot time series graph of multiple projects in data from merge_commit_ts()
plot_multiple_ts <- function(merged_ts) {
    ggplot(data=merged_ts, aes(x=t,y=ncommits,group=project,color=project)) + geom_point() + geom_line() +
      xlab("Months") +
      ylab("Number of Commits")
}
```

# Distribution of Project Duration

Let us look at how many months of data we have for each project in our dataset.
```{r, echo=FALSE, fig.height=6, fig.width=10}
g <- ggplot(data=pcommits) + 
     geom_histogram(aes(x=nmonths), color="black", fill="white", binwidth=1) +
     xlab("Number of Months") +
     ylab("Number of Projects")
g
```
Our data is heavy tailed to the right, so let's examine the data as a box plot.
```{r, echo=FALSE, fig.height=2, fig.width=10}
g <- ggplot(data=pcommits) + 
     geom_boxplot(aes(x=nmonths)) +
     xlab("Number of Months")
g
```
Let's examine the biggest outliers.
```{r}
pcommits %>% filter(nmonths>400)
```
There are 21 lexlib projects, all with large amounts of history.
```{r}
pcommits %>% filter(str_detect(project, '^lexlib_'))
```
When checking out the repository with the second longest duration according to our data, https://github.com/ronnywang/twlaw, git log shows commits only from Nov 2016 to Sep 2020, which is a much shorter time period than 458 months. On the other hand, the brlcad mirror repository, https://github.com/earl-ducaine/brlcad-mirror, had commits from Dec 1983 to Dec 2019, which is 428 months. I couldn't definitely identify the lexlib project, which has the longest duration at 506 months.

# Distribution of Project Changepoints

Let us look at how many months of data we have for each project in our dataset.
```{r, echo=FALSE, fig.height=6, fig.width=10}
g <- ggplot(data=pchangepoints) + 
     geom_histogram(aes(x=nchangepoints), color="black", fill="white", binwidth=1) +
     xlab("Number of Changepoints") +
     ylab("Number of Projects")
g
```

# A sample of projects: 40 segments of history

```{r}
fouryears <- pcommits %>% filter(nmonths==40)
fouryears
```
## Computing Change Points for the Samples

```{r}

```


## Comparing Time Series of Different Projects

```{r, echo=FALSE, fig.height=6, fig.width=12}
ts1 <- read_commit_ts(path(commits_dir, 'ALTUN69_PHP-Fusion.csv'))
ts2 <- read_commit_ts(path(commits_dir, 'AndreasSummer_Rubberduck.csv'))
mts1 <- merge_commit_ts(ts1, ts2)
plot_multiple_ts(mts1)
```

```{r, echo=FALSE, fig.height=6, fig.width=12}
ts1 <- read_commit_ts(path(commits_dir, 'ALTUN69_PHP-Fusion.csv'))
ts2 <- read_commit_ts(path(commits_dir, 'AndreasSummer_Rubberduck.csv'))
ts3 <- read_commit_ts(path(commits_dir, '01000101_cloudify-manager.csv'))
mts2 <- merge_commit_ts(mts1, ts3)
plot_multiple_ts(mts2)
```
