---
title: "Changepoint Detection for Random Sample of WoC S Time Series"
author: "James Walden"
date: "1/13/2021"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(changepoint)
library(changepoint.np)
library(fs)
library(lubridate)
library(tidyverse)
```

# Pick a random sample of projects

We create a dataset of 10 randomly chosen projects from the version S dataset from World of Code. The samples are stored in a list of tibbles named `sample_projects`.

```{r random_sample_params}
nprojects <- 10
set.seed(1)
```

```{r read_commit_ts_fn}
read_commit_ts <- function(path) {
    colnames <- c('year_month','ncommits','nauthors','timestamp')
    commits_ts <- read_csv(file=path, col_names=colnames, col_types='ciii', skip=1)
    commits_ts <- commits_ts %>%
        mutate(date=as_date(str_c(year_month, '-01')), timestamp=NULL) %>%
        complete(date=seq.Date(min(date), max(date), by="month"), fill=list(ncommits=0, nauthors=0)) %>%
        mutate(year_month=str_c(year(date),str_pad(month(date),2,"left","0"),sep='-'))
    commits_ts
}
```

```{r random_sample}
indir <- '../data/commits/auth-and-comm_data_S'
in_paths <- dir_ls(indir, glob='*.csv')
sample_paths <- sample(in_paths, nprojects)
sample_projects <- lapply(sample_paths, read_commit_ts)
s1 <- sample_projects[[1]]
```

# Examining the first project in detail

We look at changepoints in multiple time series for the first project, including number of commits and number of unique authors per month.

```{r, echo=FALSE, fig.height=4, fig.width=8}
g <- ggplot(s1, aes(x=date, y=ncommits)) + geom_point() + geom_line(group=1) + xlab("Month") + ylab("Number of Commits")
g
```

## Number of Commits: AMOC

We find a changepoint for the number of commits time series using the defaults in R changepoint, which uses the AMOC (At Most One Changepoint) algorithm.
```{r}
s1c.cpt <- cpt.mean(s1$ncommits)
s1c.pts <- cpts(s1c.cpt)
s1c.pts
```

```{r}
plot(s1c.cpt)
```

```{r}
s1c.date <- s1 %>% slice(s1c.pts) %>% select(date) %>% pull()
s1 %>% slice(s1c.pts)
```

```{r}
g + geom_vline(xintercept=s1c.date, color="blue", linetype="dashed") + 
    ggtitle("Changepoints in sample 1 commits time series")
```

## Number of Commits: Binary Segmentation

In this section, we use the binary segmentation algorithm, which can find multiple changepoints in a time series.
```{r}
s1c.binseg.cpt <- cpt.mean(s1$ncommits, method="BinSeg")
s1c.binseg.pts <- cpts(s1c.binseg.cpt)
s1c.binseg.pts
```

```{r}
plot(s1c.binseg.cpt)
```

```{r}
s1c.binseg.date <- s1 %>% slice(s1c.binseg.pts) %>% select(date) %>% pull()
s1 %>% slice(s1c.binseg.pts)
```

```{r}
g + geom_vline(xintercept=s1c.binseg.date, color="blue", linetype="dashed") + 
    ggtitle("Changepoints in sample 1 commits time series")
```

## Number of Commits: Nonparametric

We next use the nonparametric variant `changepoint.np` which finds 7 changepoints instead of one.
```{r}
s1c.cpt.np <- cpt.np(s1$ncommits)
s1commits.np <- cpts(s1c.cpt.np)
s1commits.np
```

```{r}
plot(s1c.cpt.np)
```

## Number of Authors

Now we look at the time series for number of unique authors who contributed each month.

```{r, echo=FALSE, fig.height=4, fig.width=8}
g <- ggplot(s1, aes(x=date, y=nauthors)) + geom_point() + geom_line(group=1) + xlab("Month") + ylab("Number of Authors")
g
```

We find one changepoint for number of commits using the defaults in R changepoint.
```{r}
s1c.cpt <- cpt.mean(s1$ncommits)
s1authors <- cpts(s1c.cpt)
s1authors
```
```{r}
plot(s1c.cpt)
```

```{r}
s1c.date <- s1 %>% slice(s1authors) %>% select(date) %>% pull()
s1 %>% slice(s1authors)
```

```{r}
g + geom_vline(xintercept=s1c.date, color="blue", linetype="dashed") + 
    ggtitle("Changepoints in sample 1 authors time series")
```

We next use the nonparametric variant `changepoint.np` which finds 5 changepoints instead of one.
```{r}
s1a.cpt.np <- cpt.np(s1$ncommits)
s1authors.np <- cpts(s1a.cpt.np)
s1authors.np
```

```{r}
plot(s1a.cpt.np)
```

# Finding the Number of Changepoints with CROPS

Choosing the best penalty value to find the correct number of changepoints is difficult. We have been using default penalties so far. The CROPS (Changepoints for a Range of Penalties) algorithm lets you specify minimum and maximum penalty values. It returns optimal segmentations for any penalty between these values.

## First Sample: Number of Commits

We find optimal segmentations for 1-3, 5-6, and 8 changepoints.
```{r}
s1.commits.crops <- cpt.var(s1$ncommits, method="PELT", penalty="CROPS", pen.value=c(5,500))
cpts.full(s1.commits.crops)
```

Changepoint provides a diagnostic elbow plot for CROPS, showing how many changepoints for each penalty. The intuition is that if a true changepoint is added to the model, there will be a large improvement in the test statistic. False changepoints will not improve the fit much. Therefore, you look for an elbow in the plotted curve.

For the commits time series, it looks like the optimal number of changepoints is 2.
```{r}
plot(s1.commits.crops, diagnostic=TRUE)
```

```{r}
plot(s1.commits.crops, ncpts=2)
```
Let's look at the range of penalty values that introduce new changepoints.
```{r}
pen.value.full(s1.commits.crops)
```

## First Sample: Number of Authors

We use CROPS to find the appropriate number of changepoints for the authors time series for the same project.

CROPS finds optimal segmentations for 2 and 4 changepoints.
```{r}
s1.authors.crops <- cpt.var(s1$nauthors, method="PELT", penalty="CROPS", pen.value=c(5,500))
cpts.full(s1.authors.crops)
```

The curve shows substantial improvement as changepoints are added, so it appears that the best number of changepoints is 4.
```{r}
plot(s1.authors.crops, diagnostic=TRUE)
```

```{r}
plot(s1.authors.crops, ncpts=4)
```

Let's look at the range of penalty values that introduce new changepoints.
```{r}
pen.value.full(s1.authors.crops)
```
