---
title: "Segmented Regression for Open Source Monthly Time Series"
author: "James Walden"
date: "11/22/2020"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(segmented)
library(tidyverse)
```

## Data

I'm using OpenSSL data from my MSR'20 paper, which includes time series data from project inception to the end of 2019.

I select data from 2012-2016 to analyze and call it commits for ease of typing.

```{r load_data}
openssl_commits <- read_csv('openssl-commits-per-month.csv', col_types="Di")
openssl_nearhb_commits <- openssl_commits %>% filter(Date >= as.Date('2012-01-01') & Date <= as.Date('2016-12-31'))
commits <- openssl_nearhb_commits
commits
```
## Visualizing the Time Series

```{r, echo=FALSE, fig.height=6, fig.width=10}
g <- ggplot(commits, aes(x=Date, y=ncommits)) + geom_point() + geom_line(group=1) + xlab("Month") + ylab("Number of Commits")
g
```

## Linear Model of Commits

This section shows that linear regression cannot accurately model the time series from start to finish.

Linear models handle dates by converting them to integers via R's as.numeric() function. These numbers represent the number of days since January 1, 1970.

```{r}
linmodel <- lm(ncommits ~ Date, data = commits)
summary(linmodel)
```
```{r}
lincoefs <- coef(linmodel) # linear regression intercept and slope coefficients
g + geom_abline(intercept = lincoefs[1], slope = lincoefs[2]) +
    ggtitle("Linear Regression Model of OpenSSL Commits 2012-2016")
```

## Segmented Regression

```{r}
segmodel <- segmented.lm(linmodel, seg.Z = ~ Date)
summary(segmodel)
```
We can obtain just the breakpoints by subsetting the model object.
```{r}
segmodel$psi
as.Date(segmodel$psi[2], origin=lubridate::origin)
```
Segmented started its process of finding breakpoints by guessing an initial value, which is June 16, 2014.

We can specify our own guesses with the "psi" argument to segmented().
```{r}
as.Date(segmodel$psi[1], origin=lubridate::origin)
```
By default, the segmented model found only a single breakpoint in the time series.

Add predicted values from segmented regression model to commits data frame, so we can plot actual against predicted values.
```{r}
commits$npred <- predict.segmented(segmodel)
ggplot(commits, aes(x=Date)) + 
  geom_line(aes(y = ncommits), color = "black") + 
  geom_line(aes(y = npred), color="blue", linetype="twodash") +
  ggtitle("Segmented Regression Model of OpenSSL Commits 2012-2016")
```
Let's create segmented models with multiple breakpoints.
```{r}
seg2 <- segmented.lm(linmodel, seg.Z = ~Date, npsi = 2)
summary(seg2)
```

```{r}
commits$npred2 <- predict.segmented(seg2)
ggplot(commits, aes(x=Date)) + 
  geom_line(aes(y = ncommits), color = "black") + 
  geom_line(aes(y = npred2), color="blue", linetype="twodash") +
  ggtitle("Segmented Regression Model with 2 Breakpoints")
```
```{r}
seg3 <- segmented.lm(linmodel, seg.Z = ~Date, npsi = 3)
summary(seg3)
```

```{r}
commits$npred3 <- predict.segmented(seg3)
ggplot(commits, aes(x=Date)) + 
  geom_line(aes(y = ncommits), color = "black") + 
  geom_line(aes(y = npred3), color="blue", linetype="twodash") +
  ggtitle("Segmented Regression Model with 3 Breakpoints")
```

# Will segmented always find a breakpoint?

Let's investigate how segmented handles a series with no breakpoints, because it's a horizontal line.

```{r}
nobp <- tibble(Date=commits$Date, n=1)
nobp %>% head(5)
```
```{r}
nobplm <- lm(n ~ Date, data=nobp)
nobpseg <- segmented(nobplm, seg.Z = ~Date)
summary(nobpseg)
```

Segmented finds a breakpoint, though the magnitude of the coefficient is about 10^-16 so it has almost no effect.

Let's find the date of the nonexistent breakpoint.
```{r}
nobpseg$psi
as.Date(nobpseg$psi[1], origin=lubridate::origin)
```

The magnitude of the U1.Date coefficient is about 10^-16 so it has no visible effect on the line.
```{r}
nobp$npred <- predict.segmented(nobpseg)
ggplot(nobp, aes(x=Date)) + 
  geom_line(aes(y = n), color = "black") + 
  geom_line(aes(y = npred), color="blue", linetype="twodash") +
  ggtitle("Segmented Regression Model of No Slope Line")
```

Let's try modeling a series that's almost a horizontal line but that has normally distributed variation around a value (10).
```{r}
nobp2 <- tibble(Date=commits$Date, n=10+rnorm(nrow(commits)))
nobp2 %>% head(5)
```

```{r}
nobplm2 <- lm(n ~ Date, data=nobp2)
nobpseg2 <- segmented(nobplm2, seg.Z = ~Date)
summary(nobpseg2)
```

The U1.Date coefficient is no longer tiny once we added noise.
```{r}
nobpseg$psi
as.Date(nobpseg$psi[2], origin=lubridate::origin)
```

The effect is still small enough to not be visible on the line.
```{r}
nobp$npred <- predict.segmented(nobpseg)
ggplot(nobp, aes(x=Date)) + 
  geom_line(aes(y = n), color = "black") + 
  geom_line(aes(y = npred), color="blue", linetype="twodash") +
  ggtitle("Segmented Regression Model of No Slope Line")
```

If we use segmented to identify the existence of breakpoints, we will have to identify a cutoff value for the magnitude of the U1.Date coefficient.
